preciseTAD: A machine learning framework for precise TAD boundary prediction at base level resolution

### 2020-08-15 Manuscript/manuscript.Rmd
address todo preprint items

---
title: "preciseTAD: A machine learning framework for precise TAD boundary prediction at base level resolution"
csl: styles.ref/genomebiology.csl
output:
  word_document:
    reference_docx: styles.doc/NIH_grant_style.docx
  pdf_document: 
bibliography: References/references.bib
---

@@

When built using Arrowhead and Peakachu ground truth boundaries at 5 kb and 10 kb, respectively, *preciseTAD* was able to predict a total of 7,846 TAD and 15,838 chromatin loop boundaries on GM12878, as well as 8,883 TAD and 11,016 chromatin loop boundaries on K562 [??? Need them in Supplementary Table 4, Excel, BED format, sheet 1 - GM12878 PTBR, sheet 2 - GM12878 PTBP, sheet 3,4 - K562. Supp tables 1, 2, 3 are OK, renumber subsequent. Write legend]. We reported less predicted TAD boundaries at 5 kb than Arrowhead on both cell lines (Table 1, Supplementary Table 3). This can be attributed to Arrowhead's inflation of called TADs, especially at higher resolutions, that, when visualized, often do not correspond to intra-chromosomal contacts featured on a contact matrix. Additionally, *preciseTAD* boundaries were visually more likely to demarcate high concentrations of intra-chromosomal contacts [??? Do not understand], as can be seen on a Hi-C contact matrix (Figure 5A). *preciseTAD* also predicted fewer chromatin loop boundaries than Peakachu (Table 1, Supplementary Table 3). This can be attributed to the fact that *preciseTAD* leverages four known drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143. This is in contrast to Peakachu, which considers only CTCF sites to mark the predicted boundaries as true positives. These results suggest that genome annotation-driven predicted boundaries may be more biologically relevant than those detected from Hi-C data alone.

@@

![**preciseTAD boundaries are enriched for known molecular drivers of 3D chromatin.** (A) The location of preciseTAD (green) vs. ARROWHEAD (blue) boundaries on GM12878 (chr 14: 50085000-50800000). The black line represents the predicted probability of each base being a TAD boundary. (B) Genome browse visualization of bigwig files for known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143. (C) Enriched heatmaps comparing the enrichment of CTCF signal around ARROWHEAD TAD boundaries and preciseTAD boundaries, and (D) around Peakachu loop boundaries and preciseTAD for the entire genome. (E) Log 2 genomic distances from boundaries to the nearest CTCF sites. (F) Venn diagrams illustrating the levels of conservation between TAD/loop boundaries for GM12878 (red) and K562 (blue) identified by preciseTAD (using ARROWHEAD; top left), ARROWHEAD (top right), preciseTAD (using PEAKACHU; bottom left), and PEAKACHU (bottom right). The median Jaccard Index (IQR) across each chromosome is provided below each Venn diagram](Figures/figure5.png)

@@

**Supplementary Table 4.** Additional performance metrics when implementing a random forest using
Arrowhead ground truth TAD boundaries. Performances are averaged across each holdout chromosome that was reserved for testing.

**Supplementary Table 5.** Additional performance metrics when implementing a random forest using
Peakachu ground truth chromatin loop boundaries. Performances are averaged across each holdout chromosome that was reserved for testing.

@@

>>>>

---
title: "preciseTAD: A machine learning framework for precise TAD boundary prediction at base level resolution"
csl: styles.ref/genomebiology.csl
output:
  pdf_document: 
  word_document:
    reference_docx: styles.doc/NIH_grant_style.docx
bibliography: References/references.bib
---

@@

When built using Arrowhead and Peakachu ground truth boundaries at 5 kb and 10 kb, respectively, *preciseTAD* was able to predict a total of 7,846 TAD and 15,838 chromatin loop boundaries on GM12878, as well as 8,883 TAD and 11,016 chromatin loop boundaries on K562 (Supplementary Table 4). We reported less predicted TAD boundaries at 5 kb than Arrowhead on both cell lines (Table 1, Supplementary Table 3). This can be attributed to Arrowhead's inflation of called TADs, especially at higher resolutions, that, when visualized, often do not correspond to intra-chromosomal contacts featured on a contact matrix (Figure 5A). *preciseTAD* also predicted fewer chromatin loop boundaries than Peakachu (Table 1, Supplementary Table 3). This can be attributed to the fact that *preciseTAD* leverages four known drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143. This is in contrast to Peakachu, which considers only CTCF sites to mark the predicted boundaries as true positives. These results suggest that genome annotation-driven predicted boundaries may be more biologically relevant than those detected from Hi-C data alone.

@@

![**preciseTAD boundaries are enriched for known molecular drivers of 3D chromatin.** (A) The location of preciseTAD (green) vs. Arrowhead (blue) boundaries on GM12878 (chr 14: 50085000-50800000). The black line represents the predicted probability of each base being a TAD boundary. (B) Genome browse visualization of bigwig files for known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143. (C) Enriched heatmaps comparing the enrichment of CTCF signal around Arrowhead TAD boundaries and preciseTAD boundaries, and (D) around Peakachu loop boundaries and preciseTAD for the entire genome. (E) Log 2 genomic distances from boundaries to the nearest CTCF sites. (F) Venn diagrams illustrating the levels of conservation between TAD/loop boundaries for GM12878 (red) and K562 (blue) identified by preciseTAD (using Arrowhead; top left), Arrowhead (top right), preciseTAD (using Peakachu; bottom left), and Peakachu (bottom right). The median Jaccard Index (IQR) across each chromosome is provided below each Venn diagram](Figures/figure5.png)

@@

**Supplementary Table 4.** GM12878(K562)_PTBR - The start and end coordinates defining the clusters of *preciseTAD* predicted TAD boundary regions (PTBRs) for each chromosome on GM12878 (K562). GM12878(k562)_TAD - The start and end coordinates defining TADs demarcated by a unique *preciseTAD* predicted TAD boundary point (PTBP) for each chromosome on GM12878 (K562).

*preciseTAD* predicted boundary regions (PTBR) and points (PTBP) for GM12878 and k562, using Arrowhead TAD boundary coordinates as ground truth. For PTBP,

<!--
**Supplementary Table 5.** Additional performance metrics when implementing a random forest using Arrowhead ground truth TAD boundaries. Performances are averaged across each holdout chromosome that was reserved for testing.

**Supplementary Table 6.** Additional performance metrics when implementing a random forest using Peakachu ground truth chromatin loop boundaries. Performances are averaged across each holdout chromosome that was reserved for testing.
-->

@@

AUC-Area Under the Curve; BA-Balance Accuracyl; bp-base pair; BroadHMM-BroadHMM cis-regulated elements; ChIP-seq-Chromatin Immunoprecipitation followed by high-throughput DNA sequencing; CI-Class Imbalance; CLARA-Clustering Large Applications ;DBSCAN-Density-Based Spatial Clustering of Applications with Noise; FN-False Negatives; FP-False Positives; Hi-C-High-througput chromosome conformation capture; HM-Histone Modifications; kb-kilobase; ML-Machine Learning; OC-Overlap Counts; OP-Overlap Percents; PAM-Partitioning Around Medoids; PTBP-preciseTAD Boundary Point; PTBR-preciseTAD Boundary Region; RF-Random Forest; RFE-Recursive Feature Elimination; ROC-Receiver Operating Characteristic; ROS-Random Over-Sampling
RUS	Random Under-Sampling; SMOTE-Synthetic Minority Over-sampling Technique; TAD-Topologically Associated Domains; TFBS-Transcription Factor Binding Sites; TN-True Negatives; TP-True Positives
